---
editor_options: 
  chunk_output_type: console
---

# Abundance vs. acoustic detections

In this script, we will explore correlations and regressions between abundance and acoustic detections across species and sites.

## Load necessary libraries  
```{r}
library(tidyverse)
library(dplyr)
library(stringr)
library(vegan)
library(scico)
library(data.table)
library(extrafont)
library(sf)
library(raster)
library(scales)
library(ggplot2)
library(ggspatial)
library(colorspace)
library(lubridate)
library(naniar)
library(ggstatsplot)
library(patchwork)
```

## Load dataframe containing point count and acoustic data
```{r}
datSubset <- read.csv("results/datSubset.csv")
```

## Creating a single dataframe prior to running statistical analysis

```{r}
## filter out species that were only detected in one approach and not in the other

## Nine species were detected only in point count data but not in the acoustic data and can be ignored for downstream analysis: Herring Gull, Northern Cardinal, American Goshawk, Swamp Sparrow, Red-bellied Woodpecker, Pine Siskin, Savannah Sparrow, Eastern Kingbird, Brown-headed Cowbird
## we will remove the above list of species from the point count data
point_count <- datSubset %>%
  filter(data_type == "point_count") %>%
  filter(!common_name %in% c("Herring Gull", 
                             "Northern Cardinal", "American Goshawk", 
                             "Swamp Sparrow", "Red-bellied Woodpecker", 
                             "Pine Siskin",
                             "Savannah Sparrow", "Eastern Kingbird", 
                             "Brown-headed Cowbird"
                             ))

## Six species were detected only in acoustic data and not in the point count data and can also be ignored for downstream analysis: Cape May Warbler, Gray Catbird, Mallard, Prairie Warbler, House Wren, Bobolink
acoustic_data <- datSubset %>%
  filter(data_type == "acoustic_data") %>%
  filter(!common_name %in% c("Cape May Warbler", "Gray Catbird", 
                             "Mallard", "Prairie Warbler", 
                             "House Wren", "Bobolink"
                             ))

# estimate abundance for each species for each site/visit combination
# ignore distance for now as we can model it in future analysis
abundance <- point_count %>%
  group_by(site_id, site_name, year, visit_number,
           scientific_name,
           common_name) %>% 
  summarise(abundance = sum(abundance)) %>%
  ungroup()

# estimate detections for each species for each site and year combination using the manual annotations of the acoustic data
detections <- acoustic_data %>%
  group_by(site_id, site_name, year, visit_number,
           scientific_name,
           common_name) %>% 
  summarise(detections = n()) %>%
  ungroup()

## sanity check to ensure that we are not including any species in either data type by mistake
pc_not_aru <- unique(anti_join(abundance[,6], 
                               detections[,6]))
aru_not_pc <- unique(anti_join(detections[,6], 
                               abundance[,6]))
## looks like we are good to join both datasets

# create a single dataframe
data <- full_join(abundance, detections) %>%
  replace_na(list(abundance = 0, detections = 0))
```

## Correlative analysis at a regional level

We run this analysis by pooling acoustic detections across sites within region. 
```{r}
## prior to running correlations, we get rid of species with very few counts of either abundance or detections within each year
## essentially, we want to filter out rarities at this stage 

## using a random cutoff of a minimum of 10 individuals in point count
## grouping by species, year (2022 or 2023) since vocal activity may vary by season and site_name (regional variation as a function of site/habitat)
spp_subset <-  data %>%
  group_by(common_name, year, site_name) %>%
  summarise(abundance = sum(abundance), detections = sum(detections)) %>%
  ungroup() %>%
  filter(abundance >=10 & detections >= 10)

# subset the data
# Only 38 species remain if we were to keep the above filters
data <- data %>%
  semi_join(spp_subset, by = c("common_name", "year", "site_name"))

# visualization
# for the first correlation, let's pool data at a regional level
grouped_data <- data %>%
  group_by(site_name, year, scientific_name, common_name) %>%
  summarise(abundance = sum(abundance),
            detections = sum(detections)) %>%
  ungroup()

# create a grouping variable for visualization
grouped_data <- grouped_data %>%
  mutate(site_year = paste(site_name, year, sep = " - ")) %>%
  mutate(site_year = factor(site_year, levels = unique(site_year)))

# visualize
fig_abund_detec_community <- grouped_ggscatterstats(
  data = grouped_data,
  y = abundance,
  x = detections,
  type = "r",
  ylab = "Abundance (point-count)",
  xlab = "Acoustic detections",
  grouping.var = site_year,
  label.var = common_name, 
  label.expression = detections > 500,
  ggplot.component = list(theme(text = element_text(family = "Century Gothic", size = 15, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 15, face = "bold"))))

ggsave(fig_abund_detec_community, filename = "figs/abundance_vs_detections_pooled_across_sites_by_year.png", width = 25, height = 12, device = png(), units = "in", dpi = 300)
dev.off() 

# inferences: patterns of correlations are reflective of the bird community across regional sites and years. We see fairly similar values of correlation between point count data and acoustic detections within a region across years. For example, HBEF between 2022 and 2023. However, this includes data that is pooled across species and one must test if such patterns emerge within species. 
```

## Species-specific correlations

Here, we run analysis correlating abundance and detections for each species across years. 

```{r}
# create a grouping variable for visualization
data <- data %>%
  mutate(site_year = paste(site_name, year, sep = " - ")) %>%
  mutate(site_year = factor(site_year, levels = unique(site_year)))

# plots to be created species by species
plots <- list()

for(i in 1:length(unique(data$common_name))){
  
# extract species scientific name
a <- unique(data$common_name)[i]
  
# subset data for plotting
for_plot <- data[data$common_name==a,]

# visualization
plots[[i]] <- grouped_ggscatterstats(
  data = for_plot,
  y = abundance,
  x = detections,
  type = "r",
  ylab = "Abundance (point-count)",
  xlab = "Acoustic detections",
  grouping.var = site_year,
  ggplot.component = list(theme(text = element_text(family = "Century Gothic", size = 15, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 15, face = "bold")))) + 
  plot_annotation(
  title = a,
  theme = theme(plot.title = element_text(family = "Century Gothic", 
                                         size = 20, 
                                         face = "bold", 
                                         hjust = 0.5)))
}

# plot and save as a single pdf
cairo_pdf(
  filename = "figs/abundance-acousticDetections-by-species-correlations.pdf",
  width = 25, height = 12,
  onefile = TRUE
)
plots
dev.off() 
```

## Regressions between abundance and acoustic detections

Here, we regress abundance and acoustic detections for each region and year.
```{r}
## We use the data object as it contains the subset of species with a minimum of 10 abundance values and 10 acoustic detections across sites and visits
## I am borrowing code from Mikula et al. 2020 below
## The data and scripts from their manuscript can be found here:
## https://osf.io/fa9ky/

# plots to be created for reach region and year 
plots <- list()

for(i in 1:length(unique(data$site_year))){
  
# subset data by unique region-year combination (eg. Acadia National Park - 2022)
subset <- unique(data$site_year)[i]
subset_data <- data[data$site_year==subset,]
n_species <- length(unique(subset_data$common_name))

# convert to data.table format
subset_data <- setDT(subset_data)

# extract t-value
subset_data[,  t_value := summary(lm(abundance~detections))$coefficients[6], by = common_name] 

# extract slope
subset_data[,  slope := lm(abundance~detections)%>% coef()%>% nth(2), by = common_name] 

# extract pearson's correlation
subset_data[,  pearson := cor(abundance,detections), by = common_name] 

# extract adjusted r squared
subset_data[,  r_sq := summary(lm(abundance~detections))$adj.r.squared, by = common_name] 

# create a column with the direction of the slope (whether it is positive or negative), which can be referred to later while plotting    
subset_data[, slope_dir := ifelse(slope >0, '+', '-')]
pos_reg <- paste("Positive regressions:",length(unique(subset_data$common_name[subset_data$slope_dir %in% c('+')])))

## visualization
plots[[i]] <- ggplot(subset_data, aes(x = detections,
                                              y = abundance)) +
  geom_point(color = "#9CC3D5",size = 1.2) +
  geom_smooth(data = subset_data, aes(group = common_name,
                                     color = slope_dir), 
              method = 'lm', se = FALSE, 
              linewidth = 0.7) +
  scale_color_manual(values=c("#1B9E77", "#D95F02")) +
  labs(x="\nAcoustic detections (from acoustic data)", 
       y="Abundance (from point count data)\n",
       title = subset,
       subtitle = paste0(n_species," ","species"),
       caption = pos_reg
       ) + 
  theme_bw() +
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")
}

# plot and save as a single pdf
cairo_pdf(
  filename = "figs/abundance-acousticDetections-region-year-regressions.pdf",
  width = 12, height = 7,
  onefile = TRUE
)
plots
dev.off()
```

## Species-specific regression plots

Here, we create species-specific regression plots by region and year
```{r}
# plots to be created for reach region and year 
plots <- list()

# regression summary
reg_summary <- data.frame()

# regressing species by species
for(i in 1:length(unique(data$common_name))){
  
# subset data by unique region-year combination (eg. Acadia National Park - 2022)
subset <- unique(data$common_name)[i]
subset_data <- data[data$common_name==subset,]

# convert to data.table format
subset_data <- setDT(subset_data)

# extract t-value
subset_data[,  t_value := summary(lm(abundance~detections))$coefficients[6], by = site_year] 

# extract slope
subset_data[,  slope := lm(abundance~detections)%>% coef()%>% nth(2), by = site_year] 

# extract pearson's correlation
subset_data[,  pearson := cor(abundance,detections), by = site_year] 

# extract adjusted r squared
subset_data[,  r_sq := summary(lm(abundance~detections))$adj.r.squared, by = site_year] 

## visualization
plots[[i]] <- ggplot(subset_data, aes(x = detections,
                                y = abundance)) +
  geom_point(color = "#9CC3D5",size = 1.2) +
  geom_smooth(aes(color = "#D95F02"), 
              method = 'lm', se = FALSE, 
              linewidth = 0.7) +
  geom_text(data = subset_data, 
            aes(label = paste("R² =", round(r_sq, 3))),
            x = -Inf, y = Inf, 
            hjust = -0.1, vjust = 1.5,
            family = "Century Gothic", size = 4,
            inherit.aes = FALSE) +
  facet_wrap(~site_year, scales = "free")  +
  labs(x="\nAcoustic detections (from acoustic data)", 
       y="Abundance (from point count data)\n",
       title = subset
       ) + 
  theme_bw() +
  theme(text = element_text(family = "Century Gothic", size = 18, face = "bold"),plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      plot.subtitle = element_text(family = "Century Gothic", 
      size = 15, face = "bold",color="#1b2838"),
      axis.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
      legend.position = "none")

# extract the slope, t_value, pearson correlation and the adjusted r square
lm_output <- subset_data %>%
  dplyr::select(common_name, site_year, t_value, slope, pearson,r_sq) %>% distinct()
reg_summary <- bind_rows(reg_summary, lm_output)
}

# plot and save as a single pdf
cairo_pdf(
  filename = "figs/abundance-acoustic-detections-by-species-regressions.pdf",
  width = 15, height = 10,
  onefile = TRUE
)
plots
dev.off()

# save regression summary data to file
write.csv(reg_summary, "results/abundance-acoustic-detections-by-species-regressions.csv",
          row.names = F)
```

